Model-based variable selection for regression with highly correlated variables.

### 2014-08-21 www/article.tex
up sum
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@810 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Sub-regressions between covariates}%\label{sectionOLS}
The classical linear regression model can be written
\begin{equation}
		\boldsymbol{Y}{|\boldsymbol{X}};\boldsymbol{\beta},\sigma_Y^2 = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon} \label{regressionsimple}
	\end{equation}
with $\boldsymbol{X}=(\boldsymbol{X}^1,\ldots,\boldsymbol{X}^d)$ the $n\times d$ matrix of $d$ predictor variables $\boldsymbol{X}^j$ ($j=1,\ldots,d$), $\boldsymbol{Y}$ the $n\times 1$ vector of the $n$ response variables and $\boldsymbol{\varepsilon} \sim \mathcal{N}_n(\boldsymbol{0},\sigma_Y^2\boldsymbol{I})$ the centered Gaussian noise of the regression with standard deviation $\sigma_Y >0$, $\boldsymbol{I}$ denoting the identity matrix of suitable dimension. The $d\times 1$ vector $\boldsymbol{\beta}$ gathers the coefficients of the regression\footnote{Usually a constant is included as one of the regressors. For example we can take $\boldsymbol{X}^1=(1,\ldots,1)'$. The corresponding element of $\boldsymbol{\beta}$ is then the intercept $\beta_1$.}, that can be estimated by Ordinary Least Squares (\textsc{ols}): %As shown in section \ref{sectionOLS}, 
	\begin{equation}
		\hat{\boldsymbol{\beta}}=\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}\boldsymbol{X}'\boldsymbol{Y} \label{eq:OLS}
	\end{equation}
	It is an unbiased estimate with variance matrix
	\begin{equation}
		\operatorname{Var}(\hat{\boldsymbol{\beta}})=\sigma_Y^2\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}. \label{eq:varOLS}
	\end{equation}
	This estimate requires the inversion of $\boldsymbol{X}'\boldsymbol{X}$ which can lead to great variance estimates, so unstable estimates, if it is ill-conditioned. Ill-conditioning increases when the number $d$ of covariates grows and/or when correlations within the covariates grow (in absolute value)  \cite{hoerl1970ridge}. At the limit, when $d>n$ and/or when some correlations are maximum, $\boldsymbol{X}'\boldsymbol{X}$ becomes singular. Note that $d$ and correlations are also two non unrelated factors since, in real applications, increasing $d$ makes the risk to obtain correlated covariates higher.

@@

\begin{hyp}\label{H2}
In addition, we make the complementary hypothesis that the responses covariates and the predictor covariates are totally disjoint: for any sub-regression $j=1,\ldots,d_{r}$, $J_{p}^j\subset J_f$ where $J_{r}=\{J_{r}^1,\ldots,J_{r}^{d_r}\}$ is set of all response covariates and $J_f=\{1,\ldots,d\} \backslash J_{r}$ is the set of all {\it non} response covariates. This new assumption allows to obtain very simple sub-regressions sequences, discarding hierarchical ones, in particular uninteresting cyclic sub-regressions. However it is not too much restrictive since any hierarchical (but non-cyclic) sequence of sub-regressions can be agglomerated into a non-hierarchical sequence of sub-regressions, even if it may implies to partially loose information through variance increase in the new non-hierarchical sub-regressions.
\end{hyp}

@@

\subsection{Marginal regression with decorrelated covariates}
The aim is now to use the model of linear sub-regressions $\boldsymbol{S}$ (that we assume to be known in this part) between some covariates of $\boldsymbol{X}$ to obtain a linear regression on $\boldsymbol{Y}$ relying on only uncorrelated variables $\boldsymbol{X}_f$.  The way to proceed is to marginalize the joint distribution of $\{(\boldsymbol{Y},\boldsymbol{X}_f) |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$ to obtain the distribution of $\{\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$ depending only on uncorrelated variables $\boldsymbol{X}_f$:
\begin{equation}\label{eq:marginal}
\mathbb{P}(\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}) = \int_{{\mathbb{R}^{d_r}}}\mathbb{P}(\boldsymbol{Y}| \boldsymbol{X}_f,\boldsymbol{X}_r,\boldsymbol{S};\boldsymbol{\beta},\sigma_Y^2) \mathbb{P}(\boldsymbol{X}_r | \boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\alpha},\boldsymbol{\sigma^2}) d\boldsymbol{X}_r.
\end{equation}
We need the following new hypothesis.

\begin{hyp}\label{H3}
We assume that all errors $\boldsymbol{\varepsilon}$ and $\boldsymbol{\varepsilon}_j$ ($j=1,\ldots,d_r$) are {\it mutually independent}. It implies in particular that conditional response covariates $\{\boldsymbol{X}^{J_{r}^j}|\boldsymbol{X}^{J_{p}^j},\boldsymbol{S};\boldsymbol{\alpha}_j,\sigma^2_j\}$, with distribution defined in (\ref{eq:SR}), are {\it mutually independent}:
\begin{equation}\label{eq:H3}
\mathbb{P}(\boldsymbol{X}_r | \boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\alpha},\boldsymbol{\sigma^2}) = \prod_{j=1}^{d_r} \mathbb{P}(\boldsymbol{X}^{J_{r}^j}|\boldsymbol{X}^{J_{p}^j},\boldsymbol{S};\boldsymbol{\alpha}_j,\sigma^2_j).
\end{equation}
\end{hyp}

@@

Decomposing $\boldsymbol{\beta}=(\boldsymbol{\beta}_r',\boldsymbol{\beta}_f')'$ by $\boldsymbol{\beta}_r=\boldsymbol{\beta}_{J_r}$ and $\boldsymbol{\beta}_f=\boldsymbol{\beta}_{J_f}$ the regression coefficients associated respectively to the response and to the free covariates, we can rewrite (\ref{regressionsimple}):
\begin{equation}
			\boldsymbol{Y}{|\boldsymbol{X},\boldsymbol{S}};\boldsymbol{\beta},\sigma_Y^2=\boldsymbol{X}_f\boldsymbol{\beta}_f+\boldsymbol{X}_r\boldsymbol{\beta}_r+\boldsymbol{\varepsilon}_Y. \label{eq:MainR}
\end{equation}
Combining now (\ref{eq:MainR}) with (\ref{eq:SR}), (\ref{eq:marginal}) and (\ref{eq:H3}), and also independence between each $\boldsymbol{\varepsilon}_j$ and $\boldsymbol{\varepsilon}$, we obtain the following closed-form for the distribution of $\{\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$:
\begin{eqnarray}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}&=&\boldsymbol{X}_f (\boldsymbol{\beta}_f+ \sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j)+  \sum_{j \in I_r}\beta_{j}\boldsymbol{\varepsilon}_j+\boldsymbol{\varepsilon}_Y \label{eq:Trueexpl} \\
	&=&\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl}
\end{eqnarray}
Consequently, we have obtained a new regression expression of $\boldsymbol{Y}$ but relying now {\it only} on uncorrelated covariates $\boldsymbol{X}_f$. This decorrelation process has also acted like a specific variable selection process-like because $\boldsymbol{X}_f \subseteq \boldsymbol{X}$. These two statements are expected to decrease the variance of further estimates of $\boldsymbol{\beta}$. However, the counterpart is twofold. First, this regression has a higher residual variance than the initial one since it is now $\sigma^{2*}_Y=\sigma^2_Y+\sum_{j \in I_r}\beta_{j}^2\sigma^2_j$ instead of $\sigma^2_Y$. Second, variable selection being equivalent to set $\hat{\boldsymbol{\beta}}_r=\boldsymbol{0}$, it implies possibly biased estimates of $\boldsymbol{\beta}_r$. As a conclusion, we are faced with a typical {\it bias-variance trade off}. We will illustrate it in the next section in the case of the {\sc ols} estimate.

@@

\paragraph{Qualitative illustration} Noting $\boldsymbol{\beta}^*=(\boldsymbol{\beta}_r^{*'},\boldsymbol{\beta}_f^{*'})'$ the vector of coefficients of the same dimension as $\boldsymbol{\beta}$ with $\boldsymbol{\beta}_r^*={\bf 0}$, (\ref{eq:modexpl}) can be rewritten
\begin{equation}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}=\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{X}_r\boldsymbol{\beta}_r^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl2}
\end{equation}
The {\sc ols} estimate of $\boldsymbol{\beta}^*$ is then given by
	\begin{equation}
		\hat{\boldsymbol{\beta}}^*_f=\left(\boldsymbol{X}_f'\boldsymbol{X}_f \right) ^{-1}\boldsymbol{X}_f'\boldsymbol{Y} \quad \mbox{and} \quad \hat{\boldsymbol{\beta}}^*_r = {\bf 0}.
	\end{equation}
As usual, the {\sc ols} estimate $\hat{\boldsymbol{\beta}}^*$ of $\boldsymbol{\beta}^*$ is unbiased but, however, contrary to $\hat{\boldsymbol{\beta}}$, it could be a biased estimate of $\boldsymbol{\beta}$ since
		\begin{equation}
			\mathbb{E}(\hat{\boldsymbol{\beta}}_f^*)=\boldsymbol{\beta}_f+\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \quad \textrm{and}\quad \mathbb{E}(\hat{\boldsymbol{\beta}}_r^*)=\boldsymbol{0}.
		\end{equation}
In return, its variance could be reduced compared to this one of $\hat{\boldsymbol{\beta}}$ given in (\ref{eq:varOLS}) as soon as values of $\sigma_j$ are small enough (it means strong correlations in sub-regressions) as we can see in the following expression
		\begin{equation}
			\operatorname{Var}(\hat{\boldsymbol{\beta}}_f^*)= (\sigma^2_Y+\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2)(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1} \quad \textrm{and} \quad\operatorname{Var}(\hat{\boldsymbol{\beta}}_r^*)= \boldsymbol{0}. \label{eq:varOLS*}
		\end{equation}
Indeed, no correlations between covariates $\boldsymbol{X}_f$ imply that the matrix $\boldsymbol{X}_f' \boldsymbol{X}_f$ could be sufficiently better conditioned than the matrix $\boldsymbol{X}' \boldsymbol{X}$ involved in (\ref{eq:varOLS}) to balance the added variance $\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2$ in (\ref{eq:varOLS*}). This bias-variance trade off can be resumed by the Mean Squared Error (\textsc{mse}) associated to both estimates:
	\begin{eqnarray}
			\textsc{mse}(\hat{\boldsymbol{\beta}})&=& \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}),
			 \\
			\textsc{mse}(\hat{\boldsymbol{\beta}}^*)&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}).
	\end{eqnarray}

@@

\paragraph{Numerical illustration} We now illustrate the bias-variance trade off, through a numerical example. Let a simple case with $d=5$ covariates $\boldsymbol{X}=(\boldsymbol{X}^1,\boldsymbol{X}^2, \boldsymbol{X}^4, \boldsymbol{X}^5)$ independently drawn by four independent Gaussian $\mathcal{N}_n(\boldsymbol{0},\boldsymbol{I})$. Thus, $J_f=\{1,2,4,5\}$ and $\boldsymbol{X}_f=(\boldsymbol{X}^1,\boldsymbol{X}^2,\boldsymbol{X}^4,\boldsymbol{X}^5)$. Let also $d_r=1$ response covariate $\boldsymbol{X}^3|\boldsymbol{X}^1,\boldsymbol{X}^2=\boldsymbol{X}^1+\boldsymbol{X}^2+\boldsymbol{\varepsilon}_1$ where $\boldsymbol{\varepsilon}_1\sim{\mathcal{N}_n(\boldsymbol{0},\sigma_1^2\boldsymbol{I})}$. Thus, $\boldsymbol{\alpha}_1=(1,1)'$, $\boldsymbol{J}_r=(3)$, $J_r=\{3\}$, $\boldsymbol{X}_r=(\boldsymbol{X}^3)$, $\boldsymbol{d}_p=(2)$, $\boldsymbol{J}_p=(\{1,2\})$, $\boldsymbol{X}^{\boldsymbol{J}_p^1}=(\boldsymbol{X}^1,\boldsymbol{X}^2)$ and $\boldsymbol{S}=((3),(\{1,2\}))$. Concerning now the regression with $\boldsymbol{Y}$, we define $\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{X}^1+\boldsymbol{X}^2+\boldsymbol{X}^3+\boldsymbol{X}^4+\boldsymbol{X}^5+\boldsymbol{\varepsilon}_Y$, where $\boldsymbol{\beta}=(1,1,1,1,1)'$ and $\sigma_Y \in \{10,20\}$. Finally, we deduce that $\boldsymbol{Y}|\boldsymbol{X}_f,\boldsymbol{S}=2\boldsymbol{X}^1+2\boldsymbol{X}^2+\boldsymbol{X}^4+\boldsymbol{X}^5+\boldsymbol{\varepsilon}_1+\boldsymbol{\varepsilon}_Y$
It is clear that $\boldsymbol{X}'\boldsymbol{X}$ will become more ill-conditioned as $\sigma_1$ gets smaller.

@@

\begin{figure}[h!]
	\includegraphics[width=500px]{figures/MQEexplOLSp5.png}\label{MQE1}
	\caption{Values of $\textsc{mse}(\hat{\boldsymbol{\beta}})$ (plain) and of $\textsc{mse}(\hat{\boldsymbol{\beta}}^*)$ (dotted) when varying the strength $(1-R^2)$ of the sub-regression, and also the values $n$ and $\sigma_Y$.}
\end{figure}
It appears that, when the sub-regression is strong (low $1-R^2$), $\hat{\boldsymbol{\beta}}^*$ is a better estimate than $\hat{\boldsymbol{\beta}}$: the gain on {\sc mse} can even be very significant. This effect amplified by the $\sigma_Y$ increase but is reduced by the $n$ increases. Thus, the estimate $\hat{\boldsymbol{\beta}}^*$ should be particularly useful when some covariates are highly correlated, when also the sample sizes is small and when the residual variance of $\boldsymbol{Y}$ is large. It corresponds to expected difficult practical situations.

@@

The question we address now is twofold: which criterion to retain for selecting a sub-regression structure $\boldsymbol{S}$ and which strategy to adopt for exploring the large space of models $\boldsymbol{S}$. Obviously, $\boldsymbol{S}$ being very simply understood even by non statisticians, practitioners could easily transform their expert knowledge on the phenomenon at hand, if any, into a given structure $\boldsymbol{S}$. For instance, Structural Equations Models (\textsc{sem}), which are related to our model as already mentioned in Section~\ref{sec:intro}, are often used in social sciences and economy where a structure $\boldsymbol{S}$ is generally ``hand-made''. However, in the general situation, $\boldsymbol{S}$ has to estimated.  A standard method as graphical {\sc lasso} \cite{friedman2008sparse}, which searches for a structure on the precision matrix (inverse of the variance-covariance matrix) by setting some coefficients of the precision matrix to zero, can not be applied since it is not designed to estimate oriented structures like $\boldsymbol{S}$.

>>>>

\subsection{Sub-regressions between covariates}%\label{sectionOLS}
The classical linear regression model can be written
\begin{equation}
		\boldsymbol{Y}{|\boldsymbol{X}};\boldsymbol{\beta},\sigma_Y^2 = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}_Y \label{regressionsimple}
	\end{equation}
with $\boldsymbol{X}=(\boldsymbol{X}^1,\ldots,\boldsymbol{X}^d)$ the $n\times d$ matrix of $d$ predictor variables $\boldsymbol{X}^j$ ($j=1,\ldots,d$), $\boldsymbol{Y}$ the $n\times 1$ vector of the $n$ response variables and $\boldsymbol{\varepsilon}_Y \sim \mathcal{N}_n(\boldsymbol{0},\sigma_Y^2\boldsymbol{I})$ the centered Gaussian noise of the regression with standard deviation $\sigma_Y >0$, $\boldsymbol{I}$ denoting the identity matrix of suitable dimension. The $d\times 1$ vector $\boldsymbol{\beta}$ gathers the coefficients of the regression\footnote{Usually a constant is included as one of the regressors. For example we can take $\boldsymbol{X}^1=(1,\ldots,1)'$. The corresponding element of $\boldsymbol{\beta}$ is then the intercept $\beta_1$.}, that can be estimated by Ordinary Least Squares (\textsc{ols}): %As shown in section \ref{sectionOLS}, 
	\begin{equation}
		\hat{\boldsymbol{\beta}}=\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}\boldsymbol{X}'\boldsymbol{Y} \label{eq:OLS}
	\end{equation}
	It is an unbiased estimate with variance matrix
	\begin{equation}
		\operatorname{Var}(\hat{\boldsymbol{\beta}})=\sigma_Y^2\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}. \label{eq:varOLS}
	\end{equation}
	This estimate requires the inversion of $\boldsymbol{X}'\boldsymbol{X}$ which can lead to great variance estimates, so unstable estimates, if it is ill-conditioned. Ill-conditioning increases when the number $d$ of covariates grows and/or when correlations within the covariates grow (in absolute value)  \cite{hoerl1970ridge}. At the limit, when $d>n$ and/or when some correlations are maximum, $\boldsymbol{X}'\boldsymbol{X}$ becomes singular. Note that $d$ and correlations are also two non unrelated factors since, in real applications, increasing $d$ makes the risk to obtain correlated covariates higher.

@@

\begin{hyp}\label{H2}
In addition, we make the complementary hypothesis that the response covariates and the predictor covariates are totally disjoint: for any sub-regression $j=1,\ldots,d_{r}$, $J_{p}^j\subset J_f$ where $J_{r}=\{J_{r}^1,\ldots,J_{r}^{d_r}\}$ is set of all response covariates and $J_f=\{1,\ldots,d\} \backslash J_{r}$ is the set of all {\it non} response covariates. This new assumption allows to obtain very simple sub-regressions sequences, discarding hierarchical ones, in particular uninteresting cyclic sub-regressions. However it is not too much restrictive since any hierarchical (but non-cyclic) sequence of sub-regressions can be agglomerated into a non-hierarchical sequence of sub-regressions, even if it may implies to partially loose information through variance increase in the new non-hierarchical sub-regressions.
\end{hyp}

@@

\subsection{Marginal regression with decorrelated covariates}
The aim is now to use the model of linear sub-regressions $\boldsymbol{S}$ (that we assume to be known in this part) between some covariates of $\boldsymbol{X}$ to obtain a linear regression on $\boldsymbol{Y}$ relying only on uncorrelated variables $\boldsymbol{X}_f$.  The way to proceed is to marginalize the joint distribution of $\{(\boldsymbol{Y},\boldsymbol{X}_f) |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$ to obtain the distribution of $\{\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$ depending only on uncorrelated variables $\boldsymbol{X}_f$:
\begin{equation}\label{eq:marginal}
\mathbb{P}(\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}) = \int_{{\mathbb{R}^{d_r}}}\mathbb{P}(\boldsymbol{Y}| \boldsymbol{X}_f,\boldsymbol{X}_r,\boldsymbol{S};\boldsymbol{\beta},\sigma_Y^2) \mathbb{P}(\boldsymbol{X}_r | \boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\alpha},\boldsymbol{\sigma^2}) d\boldsymbol{X}_r.
\end{equation}
We need the following new hypothesis.

\begin{hyp}\label{H3}
We assume that all errors $\boldsymbol{\varepsilon}_Y$ and $\boldsymbol{\varepsilon}_j$ ($j=1,\ldots,d_r$) are {\it mutually independent}. It implies in particular that conditional response covariates $\{\boldsymbol{X}^{J_{r}^j}|\boldsymbol{X}^{J_{p}^j},\boldsymbol{S};\boldsymbol{\alpha}_j,\sigma^2_j\}$, with distribution defined in (\ref{eq:SR}), are {\it mutually independent}:
\begin{equation}\label{eq:H3}
\mathbb{P}(\boldsymbol{X}_r | \boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\alpha},\boldsymbol{\sigma^2}) = \prod_{j=1}^{d_r} \mathbb{P}(\boldsymbol{X}^{J_{r}^j}|\boldsymbol{X}^{J_{p}^j},\boldsymbol{S};\boldsymbol{\alpha}_j,\sigma^2_j).
\end{equation}
\end{hyp}

@@

Decomposing $\boldsymbol{\beta}=(\boldsymbol{\beta}_r',\boldsymbol{\beta}_f')'$ by $\boldsymbol{\beta}_r=\boldsymbol{\beta}_{J_r}$ and $\boldsymbol{\beta}_f=\boldsymbol{\beta}_{J_f}$ the regression coefficients associated respectively to the responses and to the free covariates, we can rewrite (\ref{regressionsimple}):
\begin{equation}
			\boldsymbol{Y}{|\boldsymbol{X},\boldsymbol{S}};\boldsymbol{\beta},\sigma_Y^2=\boldsymbol{X}_f\boldsymbol{\beta}_f+\boldsymbol{X}_r\boldsymbol{\beta}_r+\boldsymbol{\varepsilon}_Y. \label{eq:MainR}
\end{equation}
Combining now (\ref{eq:MainR}) with (\ref{eq:SR}), (\ref{eq:marginal}) and (\ref{eq:H3}), and also independence between each $\boldsymbol{\varepsilon}_j$ and $\boldsymbol{\varepsilon}_Y$, we obtain the following closed-form for the distribution of $\{\boldsymbol{Y} |\boldsymbol{X}_f,\boldsymbol{S};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}\}$:
\begin{eqnarray}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}&=&\boldsymbol{X}_f (\boldsymbol{\beta}_f+ \sum_{j =1}^{d_r}\beta_{J_r^j}\boldsymbol{\alpha}_j)+  \sum_{j =1}^{d_r}\beta_{J_r^j}\boldsymbol{\varepsilon}_j+\boldsymbol{\varepsilon}_Y \label{eq:Trueexpl} \\
	&=&\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl}
\end{eqnarray}
Consequently, we have obtained a new regression expression of $\boldsymbol{Y}$ but relying now {\it only} on uncorrelated covariates $\boldsymbol{X}_f$. This decorrelation process has also acted like a specific variable selection process because $\boldsymbol{X}_f \subseteq \boldsymbol{X}$. These two statements are expected to decrease the variance of further estimates of $\boldsymbol{\beta}$. However, the counterpart is twofold. First, this regression has a higher residual variance than the initial one since it is now $\sigma^{2*}_Y=\sigma^2_Y+\sum_{j =1}^{d_r}\beta_{J_r^j}^2\sigma^2_j$ instead of $\sigma^2_Y$. Second, variable selection being equivalent to set $\hat{\boldsymbol{\beta}}_r=\boldsymbol{0}$, it implies possibly biased estimates of $\boldsymbol{\beta}_r$. As a conclusion, we are faced with a typical {\it bias-variance trade off}. We will illustrate it in the next section in the case of the {\sc ols} estimate.

@@

\paragraph{Qualitative illustration} Noting $\boldsymbol{\beta}^*=(\boldsymbol{\beta}_r^{*'},\boldsymbol{\beta}_f^{*'})'$ the vector of coefficients of the same dimension as $\boldsymbol{\beta}$ with $\boldsymbol{\beta}_r^*={\bf 0}$, (\ref{eq:modexpl}) can be rewritten
\begin{equation}
	\boldsymbol{Y}{|\boldsymbol{X}_f,\boldsymbol{S}};\boldsymbol{\beta},\boldsymbol{\alpha},\sigma_Y^2,\boldsymbol{\sigma^2}=\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{X}_r\boldsymbol{\beta}_r^*+\boldsymbol{\varepsilon}_Y^*.\label{eq:modexpl2}
\end{equation}
The {\sc ols} estimate of $\boldsymbol{\beta}^*$ is then given by
	\begin{equation}
		\hat{\boldsymbol{\beta}}^*_f=\left(\boldsymbol{X}_f'\boldsymbol{X}_f \right) ^{-1}\boldsymbol{X}_f'\boldsymbol{Y} \quad \mbox{and} \quad \hat{\boldsymbol{\beta}}^*_r = {\bf 0}.
	\end{equation}
As usual, the {\sc ols} estimate $\hat{\boldsymbol{\beta}}^*$ of $\boldsymbol{\beta}^*$ is unbiased but, however, contrary to $\hat{\boldsymbol{\beta}}$, it could be a biased estimate of $\boldsymbol{\beta}$ since
		\begin{equation}
			\mathbb{E}(\hat{\boldsymbol{\beta}}_f^*)=\boldsymbol{\beta}_f+\sum_{j =1}^{d_r}\beta_{J_r^j}\boldsymbol{\alpha}_j \quad \textrm{and}\quad \mathbb{E}(\hat{\boldsymbol{\beta}}_r^*)=\boldsymbol{0}.
		\end{equation}
In return, its variance could be reduced compared to this one of $\hat{\boldsymbol{\beta}}$ given in (\ref{eq:varOLS}) as soon as values of $\sigma_j$ are small enough (it means strong correlations in sub-regressions) as we can see in the following expression
		\begin{equation}
			\operatorname{Var}(\hat{\boldsymbol{\beta}}_f^*)= (\sigma^2_Y+\sum_{j =1}^{d_r}\sigma^2_{j}\beta_{J_r^j}^2)(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1} \quad \textrm{and} \quad\operatorname{Var}(\hat{\boldsymbol{\beta}}_r^*)= \boldsymbol{0}. \label{eq:varOLS*}
		\end{equation}
Indeed, no correlations between covariates $\boldsymbol{X}_f$ imply that the matrix $\boldsymbol{X}_f' \boldsymbol{X}_f$ could be sufficiently better conditioned than the matrix $\boldsymbol{X}' \boldsymbol{X}$ involved in (\ref{eq:varOLS}) to balance the added variance $\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2$ in (\ref{eq:varOLS*}). This bias-variance trade off can be resumed by the Mean Squared Error (\textsc{mse}) associated to both estimates:
	\begin{eqnarray}
			\textsc{mse}(\hat{\boldsymbol{\beta}})&=& \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}),
			 \\
			\textsc{mse}(\hat{\boldsymbol{\beta}}^*)&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}).
	\end{eqnarray}

@@

\paragraph{Numerical illustration} We now illustrate the bias-variance trade off, through a numerical example. Let a simple case with $d=5$ covariates $\boldsymbol{X}=(\boldsymbol{X}^1,\boldsymbol{X}^2, \boldsymbol{X}^4, \boldsymbol{X}^5)$ independently drawn by four independent Gaussian $\mathcal{N}_n(\boldsymbol{0},\boldsymbol{I})$. Thus, $J_f=\{1,2,4,5\}$ and $\boldsymbol{X}_f=(\boldsymbol{X}^1,\boldsymbol{X}^2,\boldsymbol{X}^4,\boldsymbol{X}^5)$. Let also $d_r=1$ response covariate $\boldsymbol{X}^3|\boldsymbol{X}^1,\boldsymbol{X}^2=\boldsymbol{X}^1+\boldsymbol{X}^2+\boldsymbol{\varepsilon}_1$ where $\boldsymbol{\varepsilon}_1\sim{\mathcal{N}_n(\boldsymbol{0},\sigma_1^2\boldsymbol{I})}$. Thus, $\boldsymbol{\alpha}_1=(1,1)'$, $\boldsymbol{J}_r=(3)$, $J_r=\{3\}$, $\boldsymbol{X}_r=(\boldsymbol{X}^3)$, $\boldsymbol{d}_p=(2)$, $\boldsymbol{J}_p=(\{1,2\})$, $\boldsymbol{X}^{\boldsymbol{J}_p^1}=(\boldsymbol{X}^1,\boldsymbol{X}^2)$ and $\boldsymbol{S}=((3),(\{1,2\}))$. Concerning now the regression with $\boldsymbol{Y}$, we define $\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{X}^1+\boldsymbol{X}^2+\boldsymbol{X}^3+\boldsymbol{X}^4+\boldsymbol{X}^5+\boldsymbol{\varepsilon}_Y=\boldsymbol{X}\boldsymbol{\beta}+\boldsymbol{\varepsilon}_Y$, where $\boldsymbol{\beta}=(1,1,1,1,1)'$ and $\sigma_Y \in \{10,20\}$. Finally, we deduce that $\boldsymbol{Y}|\boldsymbol{X}_f,\boldsymbol{S}=2\boldsymbol{X}^1+2\boldsymbol{X}^2+\boldsymbol{X}^4+\boldsymbol{X}^5+\boldsymbol{\varepsilon}_1+\boldsymbol{\varepsilon}_Y$
It is clear that $\boldsymbol{X}'\boldsymbol{X}$ will become more ill-conditioned as $\sigma_1$ gets smaller.

@@

\begin{figure}[h!]
	\includegraphics[width=500px]{figures/MQEexplOLSp5.png}\label{MQE1}
	\caption{Values of $\textsc{mse}(\hat{\boldsymbol{\beta}})$ (plain) and of $\textsc{mse}(\hat{\boldsymbol{\beta}}^*)$ (dotted) when varying the strength $(1-R^2)$ of the sub-regression, and also the values $n$ and $\sigma_Y$.}
\end{figure}
It appears that, when the sub-regression is strong (low $1-R^2$), $\hat{\boldsymbol{\beta}}^*$ is a better estimate than $\hat{\boldsymbol{\beta}}$: the gain on {\sc mse} can even be very significant. This effect is amplified by the $\sigma_Y$ increase but is reduced by the $n$ increases. Thus, the estimate $\hat{\boldsymbol{\beta}}^*$ should be particularly useful when some covariates are highly correlated, when also the sample sizes is small and when the residual variance of $\boldsymbol{Y}$ is large. It corresponds to expected difficult practical situations.

@@

The question we address now is twofold: which criterion to retain for selecting a sub-regression structure $\boldsymbol{S}$ and which strategy to adopt for exploring the large space of models $\boldsymbol{S}$. Obviously, $\boldsymbol{S}$ being very simply understood even by non statisticians, practitioners could easily transform their expert knowledge on the phenomenon at hand, if any, into a given structure $\boldsymbol{S}$. For instance, Structural Equations Models (\textsc{sem}), which are related to our model as already mentioned in Section~\ref{sec:intro}, are often used in social sciences and economy where a structure $\boldsymbol{S}$ is generally ``hand-made''. However, in the general situation, $\boldsymbol{S}$ has to be estimated.  A standard method as graphical {\sc lasso} \cite{friedman2008sparse}, which searches for a structure on the precision matrix (inverse of the variance-covariance matrix) by setting some coefficients of the precision matrix to zero, can not be applied since it is not designed to estimate oriented structures like $\boldsymbol{S}$.
